Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-17 Manuscript/manuscript.tex
Add version to be given to Pavel

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the \todo{Joe: I'm not sure about this: best idealization} of the ion binding as direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
The new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Ion interactions with cellular membranes play a key role in critical biological processes~\cite{seelig90,cevc90}. Ions, especially multivalent cations, modify general properties of the membrane which ultimately modulate their embedded transmembrane proteins~\cite{cevc90,tocanne90, pabst07}. Ions also have a more direct effect by modulating lipid--, protein--, and sugar--lipid interactions at the membrane surface. For example, Ca$^{2+}$  is crucial in neural signal propagation. It promotes membrane fusion by bridging lipids in the vesicles carrying neurotransmitters and the neuron synapsis membrane~\cite{REF}. Ca$^{2+}$ also participates in the T-cell receptor activation. It induces the detachment of the positively charged cytosolic tails of the CD3 protein complex from the negatively charged intracellular membrane making it accessible for Lck protein~\cite{Shi2013Ca2}. A final example involving sugar-lipid interactions mediated by ions is that Ca$^{2+}$ can modulate the presentation of the sugars found in the PI(4,5)P$_2$ lipid which ultimately modulates phospholipase C delta 1 pleckstrin homology domain (PLC $\delta$1-H)~\cite{Bilkova2017Calcium}. Interestingly, this modulation is lost when using Mg$^{2+}$ instead of Ca$^{2+}$ illustrating the selectivity of these processes towards particular ions. Despite our increasing understanding of the role of ions in cell membrane-related processes the exact molecular details of the mechanisms behind such processes remain elusive. \todo{Hector: I left one unknown citation for Pavel.}

@@

%\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.

%While relative binding affinity of different ions in PC lipids is agreed to follow Hoffmeister
%series, the molecular details of binding and binding energetics are

%The binding details, like binding sites and stoichiometry are not yet fully
%resolved but interpretation of NMR and scattering experiments suggest that one
%Ca2+ interacts mainly with the choline groups \cite{hauser76,hauser78,herbette84} of two
%phospholipid molecules \cite{altenbach84}.

@@

%Since the cation binding affinity and the headgroup 
%order parameter changes are in good agreement with experiments
%in the proposed ECC-lipids, it can be used to interpret the
%related structural changes and ion binding stoichiometry.
%New lipid models with correct ion binding affinity in lipid
%bilayers are necessary in applications of 
%MD simulations with physiological salt conditions. 
%The overestimated cation binding in the current lipid models \cite{catte16}
%may lead to significant artefacts in MD simulations. For example,
%articifially positively charged membranes overestimate interactions
%with biomolecules having opposite sign.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-POPC model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} because its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in the SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennard-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

@@

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions. \label{tab:apls}
  }
    %    { \color{red} HECTOR:Are we using here the same water model? If not is it fair the comparison considering the large changes observed in the SI table?
    %      SAMULI: For Lipid14 simulations the water is TIP3P and SPC/E for ECC-lipids. I do understand why the comparison would
    %    not be fair in this context?}
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-POPC                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is $\approx$1\AA{} smaller
%\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
%SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-POPC model vary
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

%
%Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
%to suggest a linear relation between $\alpha$-carbon order parameter
%and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
%This can be used to estimate the change in P-N vector angle
%from experiments measuring changes in $\alpha$-carbon order parameter. 
%\todo{SAMULI: I think that we should establish the relation between 
%order parameter and P-N vector angle changes by using ECC-lipids model.
%This would be useful for people measuring aplha order parameter
%changes from experiments.}
%
%For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
%addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
%gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
%\todo{I looked this roughly from the figure, should be calculated from 
%the actual relation if we decide to put this.}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    Changes of the head group order parameters of a POPC bilayer as a function of NaCl concentration
    in bulk ($C_{ion}$) from simulations with different force fields at 313 K together with 
    experimental data for DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}.
    %Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration in bulk ($C_{ion}$)
    from simulations at 313 K together with experimental data 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    %Ion concentrations in bulk water are shown in x-axis. 
    %Bulk concentrations from simulations are calculated 
    %from the farthest point from the lipid bilayer in the aqueous phase
    The error estimate for bulk concentrations was approximately 10\,mM.
    Exception was the order of magnitude larger error in
    simulation with Lipid14 and ECC-ions due to the unconverged densities  (shown if Fig.~\ref{fig:cacl-dens}) because the 
    small simulation box. 
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

@@

%\todo{HECTOR: I'm not sure I understand the meaning of "structure" in the title. In general the titles in the result section are much alike between them, specially the first ones. This makes %difficult to guess where to find a particular piece of information. I will make an effort to make them more clear.
%SAMULI: I agree. There are now new titles.}

\begin{table}[tb!]
  \caption{Bulk concentrations of Ca$^{2+}$ (C$_b$), relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$),
    and the percetages of Ca$^{2+}$ bound to phosphate or carbonyl oxyges ($r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$)
    in different POPC bilayer models. All systems have the same molar concentration of Ca$^{2+}$ with respect to water ($C_{ion}'$=350mM).
    %{\color{red}HECTOR:PLEASE CHECK}
    %{\color{red}Joe, please confirm that $r^\mathrm{Ca^{2+}} _\mathrm{PO_4}$ is really 100\% for all the Lipid14 simulations
    %  (i.e. that this is not a typo).}
  \label{tab:binding}}
  \begin{tabular}{l|c c | c | c c}
    model                  & $C_{ion}'$ & $C_{ion}\,/\,\mathrm{mM}$ & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$  & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-POPC             &  350  &  $280\pm 10 $  &  $0.06 \pm 0.01 $                           &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist     &  350  &  $210\pm 10 $  &  $0.13 \pm 0.01 $                          & 100\%  &    37\%     \\
    Lipid14/Dang           &  350  &  $160\pm 10 $  &  $0.23 \pm 0.03 $                            & 100\%  &    14\%    \\
    Lipid14/ECC-ions       &  350  &  $120\pm 100$  &  $0.35 \pm 0.11 $                         & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    %Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    %In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    The density profiles of~\ce{Cl^-} and \ce{Na^{+}} ions are divided by 2, and
    the density profiles of phosphate groups resp. water are divided by 5 resp. 200. 
    All simulations with \ce{CaCl2} shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    The simulation with \ce{NaCl} has $C_{ion}'$=1000~mM.
    }
\end{figure}

@@

\todo{HECTOR: I do fully understand the content of this paragraph. I will discuss this with Joe tomorrow. Please be carefull with my changes.} 
The histograms of residence times of \ce{Ca^{2+}} ($C_{ion}'$ = 450~mM) in a POPC bilayer for simulations using either ECC-POPC or CHARMM36 (Simulation from Refs.~\citenum{javanainen17,zenodo.259376}) are shown in Fig.~S4 in SI. In the CHARMM36 simulation, a significant number of the calciums are bound for the whole length of the trajectory (800~ns). In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the ECC-POPC model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are shorter than $60\,\mathrm{ns}$. The longest observed residence time is around 150~ns, which is well below the total length of the simulation used for analysis, i.e., 200~ns. Note that these results are in line with the experimental estimate that the residence time of \ce{Ca^{2+}} at each PC headgroup is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}. The exchange of \ce{Na+} ions at the POPC membrane is another order of magnitude faster, yielding 90\% of the residence times smaller than~1~ns, with the longest residence time being~6~ns.

%give a roughly estimate of the order
%of $1\,\mu\mathrm{s}$
%This is unfortunately not the case of many theoretical works,
%where the residence time cannot be estimated,
%or the simulation has not even decorrelated from it's 
%initial condition with a population of bound/unbound cations \cite{catte16}, 
%because it is limited by the simulation length, 
%even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
%Increasing time scales to the order of $\mu\mathrm{s}$ 
%doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used.

@@

%but we expect that the correction can be used also for other lipids
%and force fields.

%We suggest using state of the art water models like OPC3\cite{Izadi16} or OPC \cite{Izadi14},
%which yield higher accuracy than the traditional TIP3p water model~\cite{jorgensen83}.
%Several water models 
%(OPC3\cite{Izadi16}, OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}) 
%were used to exemplify the transferability of 
%the parameters of the new ECC-lipids force field.

This work can be reached as a repository containing all data at \url{zenodo.org:\dots\dots\dots}.
%and as project \mbox{NMRlipids~VI} in \url{nmrlipids.blogspot.fi}.

>>>>

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the best conception of the ion binding as direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
The new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

\section{Introduction}
Ion interactions with cellular membranes play a key role in critical biological processes~\cite{seelig90,cevc90}. Ions, especially multivalent cations, modify general properties of the membrane which ultimately modulate their embedded transmembrane proteins~\cite{cevc90,tocanne90, pabst07}. Ions also have a more direct effect by modulating lipid--, protein--, and sugar--lipid interactions at the membrane surface. For example, Ca$^{2+}$  is crucial in neural signal propagation. It promotes membrane fusion by bridging lipids in the vesicles carrying neurotransmitters and the neuron synapsis membrane~\cite{REF}. Ca$^{2+}$ also participates in the T-cell receptor activation. It induces the detachment of the positively charged cytosolic tails of the CD3 protein complex from the negatively charged intracellular membrane making it accessible for Lck protein~\cite{Shi2013Ca2}. A final example involving sugar-lipid interactions mediated by ions is that Ca$^{2+}$ can modulate the presentation of the sugars found in the PI(4,5)P$_2$ lipid which ultimately modulates phospholipase C delta 1 pleckstrin homology domain (PLC $\delta$1-H)~\cite{Bilkova2017Calcium}. Interestingly, this modulation is lost when using Mg$^{2+}$ instead of Ca$^{2+}$ illustrating the selectivity of these processes towards particular ions. Despite our increasing understanding of the role of ions in cell membrane-related processes the exact molecular details of the mechanisms behind such processes remain elusive. \todo{Hector: I left one unknown citation for Pavel.}

@@

@@

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-POPC model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} because its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in the SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennard-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.  
% SAMULI: I putted the trajectory without ions back, because it has been used as well.

@@

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions. \label{tab:apls}
  }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-POPC                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is $\approx$1\AA{} smaller
The values of the area per lipid of the ECC-POPC model vary
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.

@@

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    Changes of the head group order parameters of a POPC bilayer as a function of NaCl concentration
    in bulk ($C_{ion}$) from simulations with different force fields at 313 K together with 
    experimental data for DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}.
    %Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %Figure. in Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration in bulk ($C_{ion}$)
    from simulations at 313 K together with experimental data 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    The error estimate for bulk concentrations was approximately 10\,mM.
    Exception was the order of magnitude larger error in
    simulation with Lipid14 and ECC-ions due to the unconverged densities  (shown if Fig.~\ref{fig:cacl-dens}) because the 
    small simulation box. 
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  %Experimental values from \cite{akutsu81} for $g_3$ figure:
  %the value of $g_3$ order parameter of DPPC was -0.214 
  %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
  %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
  %at the C-2 segments of both fatty acyl chains.'' 
\end{figure}

@@

\begin{table}[tb!]
  \caption{Bulk concentrations of Ca$^{2+}$ (C$_b$), relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$),
    and the percetages of Ca$^{2+}$ bound to phosphate or carbonyl oxyges ($r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$)
    in different POPC bilayer models. All systems have the same molar concentration of Ca$^{2+}$ with respect to water ($C_{ion}'$=350mM).
  \label{tab:binding}}
  \begin{tabular}{l|c c | c | c c}
    model                  & $C_{ion}'$ & $C_{ion}\,/\,\mathrm{mM}$ & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$  & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-POPC             &  350  &  $280\pm 10 $  &  $0.06 \pm 0.01 $                           &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist     &  350  &  $210\pm 10 $  &  $0.13 \pm 0.01 $                          & 100\%  &    37\%     \\
    Lipid14/Dang           &  350  &  $160\pm 10 $  &  $0.23 \pm 0.03 $                            & 100\%  &    14\%    \\
    Lipid14/ECC-ions       &  350  &  $120\pm 100$  &  $0.35 \pm 0.11 $                         & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    %Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    the density profiles of~\ce{Cl^-} and \ce{Na^{+}} ions are divided by 2, and
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively. 
    All simulations with \ce{CaCl2} shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    The simulation with \ce{NaCl} has $C_{ion}'$=1000~mM.
    }
\end{figure}

@@

%\todo{HECTOR: I do fully understand the content of this paragraph. I will discuss this with Joe tomorrow. Please be carefull with my changes.} 
The histograms of residence times of \ce{Ca^{2+}} in a POPC bilayer ($C_{ion}'$ = 450~mM) from simulations with 
ECC-POPC and CHARMM36 (Simulation from Refs.~\citenum{javanainen17,zenodo.259376}) are shown in Fig.~S4 in SI.
In the CHARMM36 simulation, a significant number of the calciums are bound for the whole length of the trajectory (800~ns).
In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the ECC-POPC model,
where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are shorter than $60\,\mathrm{ns}$. The longest observed
residence time is around 150~ns, which is well below the total length of the simulation used for analysis, i.e., 200~ns.
Note that these results are in line with the experimental estimate that the residence time of \ce{Ca^{2+}} at each PC
headgroup is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}. The exchange of \ce{Na+} ions at the POPC membrane
is another order of magnitude faster, yielding 90\% of the residence times smaller than~1~ns, with the longest residence time being~6~ns.

@@

This work can be reached as a repository containing all data at \url{zenodo.org:\dots\dots\dots}.
